Ultrafast Control of Spin-Spin Interactions in Two-Dimensional Magnetic Systems: Kane-Mele-Hubbard Model

### 2019-03-05 Paper/paper.tex
2nd version

\textit{Introduction}.--- Discovery of athermal all-optical control of order parameters in antiferromagnetic (AFM) and ferromagnetic (FM) materials by intense and high-frequency laser pulses has elevated spintronics to a new era called ultrafast magnetism. Although the underlying microscopic mechanism behind this phenomena is unclear yet, there have been many attempts to uncover its origins. Recently, it has been theoretically proposed that laser pulses can modify and even reverse the sign of exchange interaction, the strongest spin interaction in magnetic systems. This finding has been indirectly supported by a few recent experiments. Another important spin interaction is Dzyaloshinskii-Moriya intercation (DMI), an antisymmetric exchange interaction appeared in inversion symmetry broken systems, that breaks the chiral symmetry of the system. While this interaction is much weaker than the exchange interaction, its presence has important consequences in magnetic materials such as weak ferromagnetism in AFM materials, stabilizing topological objects like chiral skyrmions and domain walls, and exotic phase of topological magnon insulators. The ratio between these two interactions control the tilting angle of canted spins. Proposing a mechanism to tune this ratio has important consequences in ultrafast spin dynamics and switching. In this Letter, we show that intense and high-frequency laser pulses can change this ratio in a class of two-dimensional (2D) magnetic materials described by the Kane-Mele-Hubbard model.

2D van der Waals magnetic materials with metallic, semiconducting and insulating band structures have been recently discovered in both AFM and FM phases. We can now use the advantage of both low dimensionality and magnetic order in the same material for the first time which promises novel spintronics devices with exceptional performances.

\textit{Model Hamiltonian}.---
Effective Hamiltonian of a 2D planar honeycomb lattice in the presence of electron-electron interactions can be described by Kane-Mele-Hubbard model \cite{Kane2005,Griset2012,Auerbach}. 
The Hamitlonian in the absence of external perturbations is the sum of the kinetic term $\hat{\mathcal{H}}_t$, the intrinsic SOI $\hat{\mathcal{H}}_{\text{SOI}}$ and electron-electron interaction modelled by extended Hubbard model $\hat{\mathcal{H}}_{\text{int}}$,
\begin{equation}
\label{MKMH}
\hat{H}_0 = \hat{\mathcal{H}}_{\text{K}} + \hat{\mathcal{H}}_{\text{SOI}} + \hat{\mathcal{H}}_{\text{int}},
\end{equation}
where
\begin{align}
\hat{\mathcal{H}}_{\text{K}} &= - \sum_{\langle i,j \rangle, \tau} t_{1}\hat{c}_{i \tau}^\dagger \hat{c}_{j \tau} - \sum_{\langle \langle i,j \rangle \rangle, \tau} t_{2}\hat{c}_{i \tau}^\dagger \hat{c}_{j \tau} \label{Ht} \\
\hat{\mathcal{H}}_{\text{SOI}} & = \sum_{\langle \langle i,j \rangle \rangle, \tau,\tau'} i\Delta\nu_{ij}\sigma^z_{\tau, \tau'}\hat{c}_{i \tau}^\dagger \hat{c}_{j \tau'} \label{Hsoi} \\
\hat{\mathcal{H}}_{\text{int}} &= U_{00}\sum_{i=1} \hat{n}_{i\uparrow}\hat{n}_{i\downarrow} + \frac{1}{2}\sum_{\langle i,j \rangle, \tau \tau'} V_{ij}\hat{n}_{i\tau}\hat{n}_{j\tau'} \label{Hint}.
\end{align}
where $\langle\cdot\rangle$ and $\langle\langle\cdot\rangle\rangle$ denote, respectively, nearest neighbors (NN) and next-nearest neighbours (NNN), $\hat{c}_{i \sigma}^\dagger$ ($ \hat{c}_{i \sigma}$) are fermionic creation (annihilation) operators for an electron on site $i$ and spin state $\tau=\{\uparrow,\downarrow\}$, $t_1$ and $t_2$ are NN and NNN hopping amplitudes, $\Delta$ is the intrinsic SOI constant, $\nu_{ij}=\pm 1$ depending on whether the electron traversing from $i$ to $j$ makes a right ($+1$) or a left turn ($-1$), $\sigma^{z}$ is the $z$-component of Pauli matrices $\bm{\sigma}$, $U_{00}$ and $V_{ij}$ are the on-site and NN Coulomb interactions.
The presence of the intrinsic NNN SOI, Eq. (\ref{Hsoi}), reduces the $\mathrm{SU}(2)$ symmetry of original Hubbard model to the $\mathrm{U}(1)$ spin group. In buckled noncoplanar honeycomb lattices and systems with structural inversion asymmetry, presence of, respectively, NNN and NN Rashba SOIs, furthur reduces the symmetry to $Z_2$.

@@

Finally, the spin Hamiltonian, at half-filling and strong interaction limit of the effective Hamiltonian $\hat{H}_{\text{eff}}$, using the relations Eqs. (\ref{SpinOperatorInv1}) and (\ref{SpinOperatorInv2}), is given by,
\begin{align}
\label{MKMHeffw}
\tilde{H}_{\text{S}}(\omega) =& \tilde{J}_{1}\sum_{\langle i,j \rangle} \bs{S}_i\cdot\bs{S}_j + \tilde{J}_{2}\sum_{\langle \langle i,j \rangle \rangle} \bs{S}_i\cdot\bs{S}_j\n \\
&+ \sum_{\langle \langle i,j \rangle \rangle} \bs{S}_i \tilde{\bs{\Gamma}}_{ij} \bs{S}_j +\sum_{\langle \langle i,j \rangle \rangle} \tilde{\bs{D}}_{ij}\cdot \bs{S}_i \times \bs{S}_j,
\end{align}
with the following renormalized spin interactions,
\begin{align}
\label{Renor-para}
\tilde{J}_{1,ij} &= 2t_1^2\frac{\mathcal{J}_{n}^2(\alpha_{ij})}{\text{U}+n\omega}, \\
\tilde{J}_{2,ij} &= 2t_2^2\frac{\mathcal{J}_{n}^2(\alpha_{ij})}{\text{U}+n\omega}, \\
\tilde{\bs{D}}_{ij} &= - 4 t_2 \Delta \frac{\mathcal{J}_{n}^2(\alpha_{ij})}{\text{U}+n\omega} \nu_{ij} \hat{\mathrm{e}}_z, \\
\tilde{\bs{\Gamma}}_{ij} &= 2\Delta^2 \text{diag}(-1,-1,1) \frac{\mathcal{J}_{n}^2(\alpha_{ij})}{\text{U}+n\omega}.
\end{align}
All spin interaction parameters are renormalized by the same function but $\alpha_{ij}$ is different for NN and NNN parameters which is peculiar of honeycomb lattices described by the Kane-Mele-Hubbard model. Thus, the ratios between these renormalized parameters are different than non-perturbed parameters. The renormalized spin interactions presented in Eq. \ref{Renor-para} are independent from helicity of laser pulse and are the same for both circular and linear polarizations.

@@

There are some recent experiments on possibility of ultrafast optical modification of exchange interactions in the bulk of iron oxides \cite{Mikhaylovskiy2015,Kim2016,Jin2014}. We hope that our work stimulate new experiments on measuring both exchange interaction and DMI in 2D magnetic systems.

>>>>

\textit{Introduction}.--- Discovery of athermal all-optical control of order parameters in antiferromagnetic (AFM) and ferromagnetic (FM) materials by intense and high-frequency laser pulses has elevated spintronics to a new era called ultrafast magnetism \cite{kimel2005ultrafast,Rasing}. Although the underlying microscopic mechanism behind this phenomena is unclear yet, there have been many attempts to uncover its origins. Recently, it has been theoretically proposed that laser pulses can modify and even reverse the sign of exchange interaction \cite{PhysRevLett.113.057201,Mentink2015,PhysRevLett.118.097701,Stepanov2017,Kitamura,MentinkReview,Barbeau2018}, the strongest spin interaction in magnetic systems. This finding has been supported by recent experiments \cite{Mikhaylovskiy2015,2018Natur.553..481G}. Another important spin interaction is Dzyaloshinskii-Moriya intercation (DMI), an antisymmetric exchange interaction appeared in inversion symmetry broken systems, that breaks the chiral symmetry of the system \cite{dzyaloshinsky1958, moriya1960,Alireza-DMI1, Alireza-DMI2}. While this interaction is much weaker than the exchange interaction, its presence has important consequences in magnetic materials such as weak ferromagnetism in AFM materials \cite{dzyaloshinsky1958, moriya1960}, stabilizing topological objects like chiral skyrmions \cite{AlirezaSk1, AlirezaSk2} and domain walls \cite{AlirezaSW}, and exotic phase of topological magnon insulators \cite{Owerre2016, Elyasi2018, Chen2018, Owerre1, Owerre2}. The ratio between these two interactions control the tilting angle of canted spins. Proposing a mechanism to tune this ratio has important consequences in ultrafast spin dynamics and switching \cite{Mikhaylovskiy2015,Mentink2015,Stepanov2017,Kitamura}. In this Letter, we show that intense and high-frequency laser pulses can change this ratio in a class of two-dimensional (2D) magnetic materials described by the Kane-Mele-Hubbard model.

2D van der Waals magnetic materials with metallic, semiconducting and insulating band structures have been recently discovered in both AFM and FM phases \cite{2D1,2D2}. We can now use the advantage of both low dimensionality and magnetic order in the same material for the first time which promises novel spintronics devices with exceptional performances.

\textit{Model Hamiltonian}.---
Effective Hamiltonian of a 2D planar honeycomb lattice in the presence of electron-electron interactions can be described by Kane-Mele-Hubbard model \cite{Kane2005,Griset2012,Auerbach}. 
The Hamitlonian in the absence of external perturbations is the sum of the kinetic term $\hat{\mathcal{H}}_t$, the intrinsic SOI $\hat{\mathcal{H}}_{\text{SOI}}$ and Coulomb interaction between electrons modelled by extended Hubbard model $\hat{\mathcal{H}}_{\text{int}}$,
\begin{equation}
\label{MKMH}
\hat{H}_0 = \hat{\mathcal{H}}_{\text{K}} + \hat{\mathcal{H}}_{\text{SOI}} + \hat{\mathcal{H}}_{\text{int}},
\end{equation}
where
\begin{align}
\hat{\mathcal{H}}_{\text{K}} &= - \sum_{\langle i,j \rangle, \tau} t_{1}\hat{c}_{i \tau}^\dagger \hat{c}_{j \tau} - \sum_{\langle \langle i,j \rangle \rangle, \tau} t_{2}\hat{c}_{i \tau}^\dagger \hat{c}_{j \tau} \label{Ht} \\
\hat{\mathcal{H}}_{\text{SOI}} & = \sum_{\langle \langle i,j \rangle \rangle, \tau,\tau'} i\Delta\nu_{ij}\sigma^z_{\tau, \tau'}\hat{c}_{i \tau}^\dagger \hat{c}_{j \tau'} \label{Hsoi} \\
\hat{\mathcal{H}}_{\text{int}} &= U_{00}\sum_{i=1} \hat{n}_{i\uparrow}\hat{n}_{i\downarrow} + \frac{1}{2}\sum_{\langle i,j \rangle, \tau \tau'} V_{ij}\hat{n}_{i\tau}\hat{n}_{j\tau'} \label{Hint}.
\end{align}
where $\langle\cdot\rangle$ and $\langle\langle\cdot\rangle\rangle$ denote, respectively, nearest neighbors (NN) and next-nearest neighbours (NNN), $\hat{c}_{i \sigma}^\dagger$ ($ \hat{c}_{i \sigma}$) are fermionic creation (annihilation) operators for an electron on site $i$ and spin state $\tau=\{\uparrow,\downarrow\}$, $t_1$ and $t_2$ are NN and NNN hopping amplitudes, $\Delta$ is the intrinsic SOI constant, $\nu_{ij}=\pm 1$ depending on whether the electron traversing from $i$ to $j$ makes a right ($+1$) or a left turn ($-1$), $\sigma^{z}$ is the $z$-component of Pauli matrices $\bm{\sigma}$, $U_{00}$ and $V_{ij}$ are the on-site and NN Coulomb interactions.
The presence of the intrinsic NNN SOI, Eq. (\ref{Hsoi}), reduces the $\mathrm{SU}(2)$ symmetry of original Hubbard model to the $\mathrm{U}(1)$ spin group. In buckled noncoplanar honeycomb lattices and systems with structural inversion asymmetry, presence of, respectively, NNN and NN Rashba SOIs, furthur reduces the symmetry to $Z_2$.

@@

Finally, the spin Hamiltonian, at half-filling and strong interaction limit of the effective Hamiltonian $\hat{H}_{\text{eff}}$, using the relations Eqs. (\ref{SpinOperatorInv1}) and (\ref{SpinOperatorInv2}), is given by,
\begin{align}
\label{MKMHeffw}
\tilde{H}_{\text{S}}(\omega) =& \tilde{J}_{1}\sum_{\langle i,j \rangle} \bs{S}_i\cdot\bs{S}_j + \tilde{J}_{2}\sum_{\langle \langle i,j \rangle \rangle} \bs{S}_i\cdot\bs{S}_j\n \\
&+ \sum_{\langle \langle i,j \rangle \rangle} \bs{S}_i \tilde{\bs{\Gamma}}_{ij} \bs{S}_j +\sum_{\langle \langle i,j \rangle \rangle} \tilde{\bs{D}}_{ij}\cdot \bs{S}_i \times \bs{S}_j,
\end{align}
with the following renormalized spin interactions,
\begin{align}
\label{Renor-para}
\tilde{J}_{1,ij} &= 2t_1^2\frac{\mathcal{J}_{n}^2(\alpha_{ij})}{\text{U}+n\omega}, \\
\tilde{J}_{2,ij} &= 2t_2^2\frac{\mathcal{J}_{n}^2(\alpha_{ij})}{\text{U}+n\omega}, \\
\tilde{\bs{D}}_{ij} &= - 4 t_2 \Delta \frac{\mathcal{J}_{n}^2(\alpha_{ij})}{\text{U}+n\omega} \nu_{ij} \hat{\mathrm{e}}_z, \\
\tilde{\bs{\Gamma}}_{ij} &= 2\Delta^2 \text{diag}(-1,-1,1) \frac{\mathcal{J}_{n}^2(\alpha_{ij})}{\text{U}+n\omega}.
\end{align}
All spin interaction parameters are renormalized by the same function but $\alpha_{ij}$ is different for NN and NNN parameters. Thus, the ratios between these renormalized parameters are different than non-perturbed parameters which is a peculiar of honeycomb lattices described by the Kane-Mele-Hubbard model. The renormalized spin interactions presented in Eq. (\ref{Renor-para}) are helicity independent.

@@

Possibility of ultrafast optical modification of exchange interactions in the bulk of iron oxides has been recently reported \cite{Mikhaylovskiy2015}. We hope that our work stimulate new experiments on measuring both exchange interaction and DMI in novel 2D magnetic systems.
